---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gghat

<!-- badges: start -->
<!-- badges: end -->

The goal of gghat is to add functionality to ggplot, such that hat graphs are included
in its repertoire. To this end, gghat contains the hat geom as well as compatible position 
functions to be used with the text and errorbar geoms. 

## Installation

You can install the development version of gghat from GitHub:

``` r
# install.packages("devtools")
# library(devtools)
# devtools::install_github("eliashyde/gghat")
```

## Example

Below is an example of gghat's functionality in the context of the tidyverse.
For a more detailed exploration of the package, please look to the relevant vignette.
This code just gives a glimpse of a few of pertinent features: 

```{r example, dpi = 300, fig.width = 10, fig.height = 6}
library(tidyverse)
library(gghat)

my_data <- tibble(
  treatment = c("A", "B", "C", "D"),
  pre_treatment_means = c(50, 54, 48, 50),
  post_treatment_means = c(52, 49, 57, 53)
) |> 
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |> 
  mutate(condition = fct_relevel(condition, "pre_treatment_means", "post_treatment_means"))

ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(backwards = FALSE, width = 0.75, alpha = 0.5) +
  geom_text(aes(label = mean), family = "Courier",
            position = position_dodgedifftext(width = 0.75, nudge = 0.6)) +
  coord_cartesian(ylim = c(44, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.1, color = "black"),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Treatment") +
  ylab("Measurement") +
  ggtitle("Measurements PRE- and POST-treatment")
```
