---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gghat

<!-- badges: start -->
<!-- badges: end -->

The goal of gghat is to add functionality to ggplot2, such that hat plots are included
in its repertoire. To this end, gghat includes geom_hat() as well as compatible positions 
to be used with geom_text() and geom_errorbar(). 

## Installation

You can install the development version of gghat from GitHub:

``` r
# install.packages("devtools")
# library devtools
devtools::install_github("eliashyde/gghat")
```

## Example

Below is an example of gghat's functionality in the context of the tidyverse:

```{r example, dpi = 300}
library(tidyverse)
library(gghat)

my_data <- tibble(
  treatment = c("A", "B", "C", "D"),
  pre_treatment_means = c(50, 53, 48, 50),
  post_treatment_means = c(51, 49, 55, 52)
) |> 
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |> 
  mutate(condition = fct_relevel(condition, "pre_treatment_means", "post_treatment_means"))

ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(backwards = FALSE, width = 0.75) +
  geom_text(aes(label = mean),
            position = position_dodgedifftext(width = 0.75, nudge = 0.65)) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  xlab("") +
  ylab("")
```
