---
title: "gghat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gghat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

In this vignette, I will explain hat graphs — what they are and from whence they came. 
I will then work through the contents of `gghat`, which allows for the creation 
of these graphs using `ggplot`. Obviously, `gghat` requires that `ggplot` is also installed. 
But I would recommend that the user installs the entire `tidyverse` too, as it contains a 
host of useful packages. The `gghat` package itself can be found on [GitHub](https://github.com/eliashyde/gghat).

```{r setup, message=FALSE}
library(gghat)
library(patchwork)
library(randomizr)
library(estimatr)
library(tidyverse)
library(vayr)
```

Comparative bar graphs — in which bars are paired or otherwise grouped — are useful because they give a strong visual impression. As always, though, we must be conscientious of how we are curating this impression. Starting the y-axis at zero has long since been the golden rule with these types of plots, lest the relative lengths of truncated bars mislead the viewer with a lie factor greater than one (Tufte, 2001). That said, a y-axis that begins at zero may minimize statistically significant treatment effects in comparative bar graphs, rendering them visually insignificant. In a recent paper, Jessica Witt suggests that we derive our y-ranges from standard deviation when possible to increase sensitivity, reducing the discrepancy between actual and perceived effect size (Witt, 2019a).

How, then, might we mitigate the negative effects of an abbreviated y-axis? We could plot only the differences between groups, allowing us to start our y-axis at zero while also limiting it to a reasonable range. In doing so, though, we would lose contextualizing information — namely, the raw values. Or instead of bars, we could plot points, which visually retain only relative position and thus do not suffer from non-zero baselines. But Witt argues that points do not cater to the principles of grouping, failing to “facilitate the perceptual grouping of pairs” (Witt, 2019b). Of course, we could connect the grouped points with lines, but this isn’t ideal either because lines imply continuous variables; they are less suited for discrete data (Zacks & Tversky, 1999).

Witt explains that comparative bar graphs need not fall by the wayside, at least not entirely; they need only be redesigned. Rather than equivocate, negotiating the aforementioned trade-offs, she presents the hat graph, a new visualization of bar-graph descent that allows for  “proper interpretation of discrete categories” and the use of  “unrestricted settings for the y-axis” (Witt, 2019b). From a pair of bars in a bar graph, a hat graph would retain both heights as horizontal lines; it would also retain a bit of the body of the second bar to represent the difference between the two heights, resulting in a hat shape. The steps are illustrated below. The “brim” of each hat marks one value (e.g., either pre-treatment or without treatment) and the “crown” marks another (e.g., either post-treatment or with treatment), while the profile reflects the difference therebetween (Witt, 2019b). As Tufte would recommend, this gets rid of redundancy. Moreover, as a graphic for difference, the profile has a lie factor of one, while the brim and crown are retained as annotations so as not to lose sight of the actual data values Hat graphs are efficient, requiring very little work on behalf of the viewer — although they are still novel, so it is to be expected that they occasion brief pause.

```{r introduction, dpi = 96, fig.width = 10, fig.height = 6, echo = FALSE}
my_data <- tibble( 
  treatment = c("A", "B", "C", "D"),
  pre_treatment_mean = c(50, 49, 48, 50),
  post_treatment_mean = c(51, 53, 55, 52)
) |>
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |> 
  mutate(condition = fct_relevel(condition, "pre_treatment_mean", "post_treatment_mean"))

plot1 <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_bar(aes(fill = condition), stat = "identity", position = "dodge", width = 0.75) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("grey50", "grey70")) +
  coord_cartesian(ylim = c(0, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5)
  ) +
  xlab("") +
  ylab("")

plot2 <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_bar(aes(fill = condition), stat = "identity", position = "dodge", width = 0.75) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("grey50", "grey70")) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5)
  ) +
  xlab("") +
  ylab("")

plot3 <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_bar(aes(fill = condition), stat = "identity", position = "dodge", width = 0.75, alpha = 0.5) +
  geom_hat(width = 0.75, fill = NA, color = "black", backwards = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("grey50", "grey70")) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5)
  ) +
  xlab("") +
  ylab("")

plot4 <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(width = 0.75, fill = "black", color = "black") +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5)
  ) +
  xlab("") +
  ylab("")

(plot1 + plot2) / (plot3 + plot4)
```

It is difficult to create hat graphs with `ggplot`. In essence, the user must “hack” a stacked and dodged bar graph — and to be clear, a stacked and dodged bar graph is itself no easy feat within current constraints. But this modest package, `gghat`, adds the necessary functionality to allow the user to quickly and easily make hat graphs. In other words, it adds hat graphs to ggplot’s repertoire. First and foremost, gghat introduces `geom_hat()`, which takes the same parameters as a pre-summarized comparative bar graph: a discrete `x` variable, a quantitative `y` variable compatible with `stat = identity`, and a `group` variable. Do note that hat graphs generally posit a pre-treatment or control group on the left, so with respect to the grouping variable users must take care to order factor levels appropriately. Then the typical aesthetic parameters apply: `fill`, `color`, `alpha`, `width`, `linewidth`, and so on. The `linewidth`, 0.75 by default, is of particular importance because it controls the weight of the hat’s brim.

`geom_hat()` is based on a position function, `dodgediff()`, which is written within the `gghat` package. That is to say, much of the work that distinguishes `geom_hat()` from, say, the `geom_bar()` or `geom_col()` is actually outsourced to this position function. The benefit of this sort of organization is that variations of the same position function can be applied to other geoms to make them compatible with hat graphs. In particular the `dodgedifftext()` position function can be used with `geom_text()`. A `nudge` argument can be passed to `dodgedifftext()`, moving text vertically away from the hats.

```{r hatplot, dpi = 96, fig.width = 10, fig.height = 6}
my_data <- tibble( # Create summarized data by treatment group, before and after
  treatment = c("A", "B", "C", "D"),
  pre_treatment_mean = c(50, 49, 48, 50),
  post_treatment_mean = c(51, 53, 55, 52)
) |>
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |> # Format as long data
  mutate(condition = fct_relevel(condition, "pre_treatment_mean", 
                                 "post_treatment_mean")) # Order properly

ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(width = 0.75, linewidth = 0.75, # Add hat geoms
           fill = "black", color = "black", alpha = 0.25) +
  geom_text(aes(label = mean), family = "Courier", # And direct labels
            position = position_dodgedifftext(width = 0.75, nudge = 0.6)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Treatment") +
  ylab("Measurement") +
  ggtitle("Measurements PRE- and POST-treatment") # An example of just one use for hat graphs
```

The `backwards` parameter, which takes a boolean argument, is unique to `gghat`. It determines the behavior of `geom_hat()` given a negative treatment effect. By default, `backwards` is equal to FALSE, such that the leftmost group is the brim of the hat, representing the baseline. Thus, a negative treatment effect would yield an upside-down hat. This is recommended, as it’s easier to intuit the underlying experimental process. But if the user prefers, they can instead set `backwards` equal to TRUE, which makes the group with the minimum value the brim, such that a negative treatment effect would be reflected by a backwards hat. In this case, if the user is applying the `dodgedifftext()` position function to `geom_text()`, they must also explicitly set `backwards` equal to TRUE within these functions to ensure proper behavior. For instance, text will sit below negative treatment effects when `backwards` is FALSE but above all groups when `backwards` is TRUE. Note that the `nudge` parameter for `dodgedifftext()` is particular in its ability to handle either case, as it pushes text above the hats upwards and text below the hats downwards. 

```{r backwards, dpi = 96, fig.width = 10, fig.height = 4}
my_data <- tibble( # New data wherein treatment B has a NEGATIVE effect
  treatment = c("A", "B", "C", "D"),
  pre_treatment_mean = c(50, 53, 48, 50),
  post_treatment_mean = c(51, 49, 55, 52) 
) |>
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |>
  mutate(condition = fct_relevel(condition, "pre_treatment_mean",
                                 "post_treatment_mean"),
         hue = if_else(treatment == "B", "YES", "NO")) # To highlight backwards hats

forwards <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(aes(fill = hue, color = hue), width = 0.75, backwards = FALSE) + # The default value
  geom_text(aes(label = mean), family = "Courier",
            position = position_dodgedifftext(width = 0.75, nudge = 0.6)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_manual(values = c("grey80", "black")) +
  scale_fill_manual(values = c("grey80", "black")) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Treatment") +
  ylab("Measurement") +
  ggtitle("Backwards = FALSE")

backwards <- ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(aes(fill = hue, color = hue), width = 0.75, backwards = TRUE) + # Backwards here
  geom_text(aes(label = mean), family = "Courier", # Backwards parameter used again below
            position = position_dodgedifftext(width = 0.75, nudge = 0.6, backwards = TRUE)) + 
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_manual(values = c("grey80", "black")) +
  scale_fill_manual(values = c("grey80", "black")) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Treatment") +
  ylab("Measurement") +
  ggtitle("Backwards = TRUE")

forwards + backwards
```

Although hat graphs are designed with paired groups in mind, `geom_hat()` can tolerate more than two groups without any issue. `backwards` can still be FALSE or TRUE per the user’s preference, and the supplementary position function `dodgedifftext()` will behave properly.

```{r multihats, dpi = 96, fig.width = 10, fig.height = 6}
my_data <- tibble(
  treatment = c("A", "B", "C", "D"),
  pre_treatment_mean = c(50, 53, 51, 50),
  post_treatment_mean_round1 = c(51, 50, 49, 52),
  post_treatment_mean_round2 = c(49, 48, 52, 53) # Add a round of treatment
) |>
  pivot_longer(!treatment, names_to = "condition", values_to = "mean") |>
  mutate(condition = fct_relevel(condition, "pre_treatment_mean", 
                                 "post_treatment_mean_round1", 
                                 "post_treatment_mean_round2")) # Always order thoughtfully

ggplot(my_data, aes(x = treatment, y = mean, group = condition)) +
  geom_hat(width = 0.75, fill = "black", color = "black", alpha = 0.25) +
  geom_text(aes(label = mean), family = "Courier",
            position = position_dodgedifftext(width = 0.75, nudge = 0.6)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(40, 60)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Courier"),
    legend.position = "none",
    panel.grid = element_blank(),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Treatment") +
  ylab("Measurement") +
  ggtitle("Measurements after MULTIPLE rounds of treatment")
```

Finally, the functionality of `gghat` is demonstrated below with a more in-depth example,
which also illustrates some of the data-level work that can go into a hat graph. Specifically,
this example uses `patriot_act` data from the `vayr` package. The Patriot Act was an anti-terrorism law, and the `patriot_act` data contains data relating to an experiment that measured support for this law after randomly exposing participants to statements that cast the legislation in either a negative or positive light. The experiment was conducted in 2009 with a nationwide sample, 
and it was replicated in 2015 with a sample of MTurkers. In both instances, the 
treatments had a similar effect on Democrats and Republicans. There are four 
variables in the data: 

  + `sample_label`, the study to which the participant belonged
  + `pid_3`, the partisanship of the participant
  + `T1_content`, the statements to which the participant was exposed
  +  `PA_support`, the participant's post-treatment support for the Patriot Act 

Below, the data is summarized and the group means are calculated. It would then be possible to calculate the standard errors of these group means and create confidence intervals from them. But better yet, because hat graphs are designed to represent the difference between control and treatment groups, confidence intervals can be created for the average treatment effect. These confidence intervals, which will rest only on the crowns of the hats not the brims, are created separately and then appended to the summarized data for plotting.

``` {r example, dpi = 96, fig.width = 10, fig.height = 6}
patriot_act <-
  patriot_act |>
  mutate(T1_content = relevel(factor(T1_content), ref = "Control"))

summary_df <-
  patriot_act |>
  group_by(T1_content, pid_3, sample_label) |>
  reframe(tidy(lm_robust(PA_support ~ 1)))

estimates_df <-
  patriot_act |>
  group_by(pid_3, sample_label) |>
  reframe(tidy(lm_robust(PA_support ~ T1_content))) |>
  filter(term!= "(Intercept)")

control_means <-
  summary_df |>
  filter(T1_content == "Control") |>
  transmute(pid_3, sample_label,
            control_mean = estimate)

estimates_df <-
  estimates_df |>
  left_join(control_means, by = join_by(pid_3, sample_label))

estimates_df  <-
  estimates_df |>
  transmute(
    pid_3,
    sample_label,
    T1_content = factor(str_remove(term, "T1_content"), levels = levels(patriot_act$T1_content)),
    conf.low_effect = conf.low + control_mean,
    conf.high_effect = conf.high + control_mean)

gg_df <-
  summary_df |> left_join(estimates_df, by = join_by(T1_content, pid_3, sample_label))

ggplot(gg_df, aes(x = pid_3, y = estimate, group = T1_content)) +
  geom_hat(fill = NA) +
  geom_linerange(aes(ymin = conf.low_effect, ymax = conf.high_effect, color = pid_3),
                 position = position_dodge(width = 0.75),
                 linewidth = 0.25, alpha = 0.75) +
  geom_text(aes(label = T1_content),
            position = position_dodgedifftext(nudge = 0.1),
            family = "Courier", fontface = "bold") +
  scale_color_manual(values = c("blue3", "red3")) +
  facet_wrap(~sample_label) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(family = "Courier"),
        panel.grid = element_blank(),
        axis.line = element_line(linewidth = 0.5),
        axis.ticks = element_line(linewidth = 0.5),
        axis.title.x = element_blank()) +
  labs(y = "Do you oppose or support the Patriot Act?\n[1: Strongly oppose to 7: Strongly support]")
```

## References

Tufte, E. R. (2001). *The Visual Display of Quantitative Information* (2nd ed.). Graphics Press.

Witt, J. K. (2019a). "Graph Construction: An Empirical Investigation on Setting the Range of the Y-Axis." *Meta-Psychology*, 3(1). https://doi.org/10.31234/osf.io/jmqkx.

—. (2019b). "Introducing Hat Graphs." *Cognitive Research*, 4(31). https://doi.org/10.1186/s41235-019-0182-3. 

Zacks, J., & Tversky, B. (1999). "Bars and Lines: A Study of Graphic Communication." *Memory & Cognition*, 27(6), 1073–1079. https://doi.org/10.3758/BF03201236.
